\bf A Unitary RG-based Auxiliary Model Approach to Strongly-Correlated Electrons

### 2025-04-11 TilingProjectDraft/tilingProject.tex
updated notes with complete HKM calculations

\documentclass[reprint,hidelinks,onecolumn]{revtex4-2}
\usepackage{braket,amsmath,color,graphicx,bookmark,bm,amssymb}
\graphicspath{{./figures/}}
\bibliographystyle{apsrev4-2}
\begin{document}

@@

\subsection{Marginal Fermi liquid behaviour in the pseudogap}\label{marginalFL}
It is well-known that the two-channel Kondo model hosts non-Fermi liquid excitations at low-energies~\cite{Tsvelick_weigmann_mchannel_1985,emery_kivelson}; that this also happens in the present model is corroborated by our calculations of the quasiparticle residue \(Z_\text{imp}\) for the local gapless excitations. For a single-particle Greens function of an interacting model, the quasiparticle residue \(Z\) is defined as the spectral weight of low-energy resonance within the spectral function:
\begin{equation}\begin{aligned}
	G_k = \frac{Z}{\omega - \varepsilon_k - i \Sigma^{\prime\prime}} + G_\text{incoh}~,\quad 1/Z = 1- \frac{d \Sigma^{\prime}}{d\omega}~,
\end{aligned}\end{equation}
where \(\Sigma^\prime\) and \(\Sigma^{\prime\prime}\) are the real and imaginary parts of the self-energy, and \(G_\text{incoh}\) is contribution to the Greens function arising from incoherent short-lived excitations as well as high-energy features that are decoupled from the low-energy excitations~\cite{varma-physrevlett.63.1996}. Small values of \(Z\) indicate that the excitations cannot be described in terms of a Fermi liquid theory. Indeed, as shown in the right panel of Fig.~\ref{channelDecoupling}, the quasiparticle residue \(Z_\text{imp}\) of the impurity excitations becomes vanishingly small in the pseudogap region, signalling that the excitations are of a non-Fermi liquid nature. Given that the pseudogap phase is close to a critical point in the impurity model phase diagram (separating the local Fermi liquid and local moment phases), it is not surprising that a quantum critical model emerges at that point.

Coleman and Ioffe have shown that by mapping the two-channel Kondo problem to a "compactified" form where the impurity spin hybridises with the spin and isospin degrees of freedom of a single conduction channel, the impurity self-energy can be shown to display marginal Fermi liquid behaviour~\cite{Coleman_tsvelik,Schofield1997}:
\begin{equation}\begin{aligned}
	\Sigma \propto \omega \ln \omega~.
\end{aligned}\end{equation}
As some of us have shown in a previous work~\cite{Patra2023MCK}, this can also be obtained from a renormalisation group calculation of the low-energy excitations of the two-channel overscreened fixed point. For the sake of completeness, we have reproduced the calculation here. By considering minimal hopping into the conduction bath, the low-energy excitations above the 2CK ground state are described by the following Hamiltonian~\cite{Patra2023MCK}:
\begin{eqnarray}
	\label{nfl_terms}
	H_\mathrm{eff} =& \frac{8t^2}{{J}}\left[\left(S^z_{0,1}\right)^2 S^+_{0,2} + \left(S^z_{0,2}\right)^2 S^+_{0,1}\right] \left(S^-_{1,1} + S^-_{1,2}\right) + \mathrm{h.c.};
\end{eqnarray}
where \({\boldsymbol S}_{i,l} = \frac{1}{2}\sum_{\sigma_1,\sigma_2}c^\dagger_{i,\sigma_1,l}{\boldsymbol \sigma}_{\sigma_1,\sigma_2}c_{i,\sigma_2,l}\) are spin operators for the \(i^\text{th}\) site on the \(l^\text{th}\) conduction channel (\(l\) can be 1 or 2 for a two-channel model). The labels \(\sigma_1\) and \(\sigma_2\) are spin indices that can be \(\pm 1\). The site index \(0\) represents the bath sites connected immediately to the impurity site. The operator \(c^\dagger_{i,\sigma,l}\) creates an electron with spin \(\sigma\) on the \(i^\text{th}\) site of the \(l^\text{th}\) conduction channel.

\subsection{Emergence of long-ranged interactions near the NFL-Mott insulator transition}
As the bath interaction \(W\) is tuned through the L-PG phase, the nodal region is the last to decouple from the impurity. This allows us to write down a simpler Kondo model near the transition, where only the nodal region is hybridising with the impurity spin through Kondo interactions. This is done by retaining only those scattering processes \({\bf k}_1 \to {\bf k}_2\) that originate from and end at \(k-\)points within a small neighborhood of width \(|{\bf q}|\) around the four nodal points: \({\bf k}_1, {\bf k}_2 \in {\bf N} + {\bf q}, |{\bf q}| \ll \pi\), where \({\bf N}\) can be any one of the four nodal points \({\bf N}_{1} = \left(\pi/2, \pi/2\right), {\bf N}_{2} = \left(-\pi/2, \pi/2\right)\) and \({\bf N}_1 + {\bf Q}_1\) and \({\bf N}_{2} + {\bf Q}_2\), where \({\bf Q}_1 = \left(-\pi,-\pi \right)\) and \({\bf Q}_2 = \left( \pi, -\pi \right) \) are the two nesting vectors. We have already seen, in Subsection ~\ref{2CKEmerge}, that the pseudogap regime displays a decoupling of the Brillouin zone into sets of points that have no explicit connecting term in the Hamiltonian. Near the transition, the neighbourhoods of \({\bf N}_{1}\) and \({\bf N}_{1} + {\bf Q}_1\) form one such connected set, while the other two nodes \({\bf N}_{2}\) and \({\bf N}_{2} + {\bf Q}_2\) form another connected set, but these two sets are disconnected from each other, in the sense of a two-channel model. We also assume that the window of \({\bf q}\) is small enough so that the fixed point Kondo coupling values \(J^*({\bf q}_1, {\bf q}_2)\) for the scattering processes involving \({\bf q}_1\) and \({\bf q}_2\) can be replaced by an average value \(J^*\).

We find that the sum degrees of freedom \(s_{{\bf q},\sigma}\) do not enter the Kondo scattering Hamiltonian \(\tilde H_\text{imp-cbath}\) (for the single channel \(l=1\)), and it is expressed purely in terms of the relative degrees of freedom:
\begin{equation}\begin{aligned}
	\tilde H_\text{imp-cbath} = J^*\sum_{{\bf q}_1, {\bf q}_2,\alpha,\beta}{\bf S}_d\cdot{\boldsymbol\sigma}_{\alpha\beta} {\phi}^\dagger_{{\bf q}_1\alpha} \phi_{{\bf q}_2\beta}~.
\end{aligned}\end{equation}
We now proceed to rewrite the other parts of the Hamiltonian in terms of these new degrees of freedom. The kinetic energy can be written as
\begin{equation}\begin{aligned}\label{kineticEnergy}
	\tilde H_\text{cbath} &= \sum_{{\bf q},\sigma}\varepsilon_{{\bf N}_1 + {\bf q}}\left(n_{{\bf N}_1 + {\bf q},\sigma} + n_{{\bf N}_1 + {\bf Q}_1 - {\bf q},\sigma}\right)\\
						  &= \sum_{{\bf q},\sigma}\epsilon_{{\bf q}}\left({\psi}^\dagger_{{\bf q},\sigma} {\phi}_{{\bf q},\sigma} + {\phi}^\dagger_{{\bf q},\sigma} \psi_{{\bf q},\sigma}\right)~,
\end{aligned}\end{equation}
where \(\epsilon_{{\bf q}}\) is a shorthand for \(\varepsilon_{{\bf N}_1 + {\bf q}}\).
We also consider a simplified form of the bath interaction (for the channel \(l=1\)), taking into account the density-density correlations in \(k-\)space:
\begin{widetext}
\begin{equation}\begin{aligned}
	\tilde H_\text{cbath-int} =& -W\sum_{{\bf q},\sigma}\left(n_{{\bf N}_1 + {\bf q},\sigma} + n_{{\bf N}_1 + {\bf Q}_1 - {\bf q},\sigma}\right) + W\sum_{{\bf q}_1,{\bf q}_2}\left(n_{{\bf N}_{1} + {\bf q}_1, \uparrow} + n_{{\bf N}_{1} + {\bf Q}_{1} - {\bf q}_1, \uparrow} \right) \left( n_{{\bf N}_{1} + {\bf q}_2, \downarrow} + n_{{\bf N}_{1} + {\bf Q}_{1} - {\bf q}_2, \downarrow} \right)~,\\
	=& -W\sum_{{\bf q},\sigma}\left(s_{{\bf q},\sigma} + r_{{\bf q},\sigma}\right) + W\sum_{{\bf q}_1,{\bf q}_2}\left(s_{{\bf q}_1,\uparrow} + r_{{\bf q}_1,\uparrow}\right) \left(s_{{\bf q}_2,\downarrow} + r_{{\bf q}_2,\downarrow}\right)~.
\end{aligned}\end{equation}
\end{widetext}
We used eq.~\ref{numberOperatorRelation} in order to obtain the final form.

We now calculate the various terms in \(\Delta \tilde H\):
\begin{equation}\begin{aligned}\label{kondoPert}
	&\mathcal{P}_L \mathcal{V} \mathcal{P}_H\frac{1}{E_L - H_D} \mathcal{V} \mathcal{P}_L = \left(\frac{1}{2} + S_d^z\right) \sum_{{\bf q}_1,{\bf q}_2}\frac{{J^*}^2}{-4W}\phi^\dagger_{{\bf q}_1,\downarrow}\phi_{{\bf q}_2, \uparrow} \left(\phi^\dagger_{{\bf q}_2,\uparrow}\phi_{{\bf q}_1, \downarrow} + (1 - \delta_{{\bf q}_1,{\bf q}_2})\phi^\dagger_{{\bf q}_1,\uparrow}\phi_{{\bf q}_2, \downarrow}\right)\\
	&= -\sum_{{\bf q}_1,{\bf q}_2}\frac{{J^*}^2}{4W}\left[r_{{\bf q}_1 \downarrow} \left(1 - r_{{\bf q}_2 \uparrow}\right) - (1 - \delta_{{\bf q}_1,{\bf q}_2})\phi^\dagger_{{\bf q}_1,\uparrow}\phi^\dagger_{{\bf q}_1,\downarrow}\phi_{{\bf q}_2, \downarrow}\phi_{{\bf q}_2, \uparrow}\right]\\
	&\mathcal{P}_L \mathcal{V} \mathcal{P}_H\frac{1}{E_L - H_D} \mathcal{V} \mathcal{P}_L = \sum_{{\bf q}_1,{\bf q}_2}\frac{{J^*}^2}{-4W}\left[r_{{\bf q}_1 \uparrow} \left(1 - r_{{\bf q}_2 \downarrow}\right) - (1 - \delta_{{\bf q}_1,{\bf q}_2})\phi^\dagger_{{\bf q}_1,\downarrow}\phi^\dagger_{{\bf q}_1,\uparrow}\phi_{{\bf q}_2, \uparrow}\phi_{{\bf q}_2, \downarrow}\right] \\
\end{aligned}\end{equation}
In both terms, we set \(S_d^z = 0\) owing to local SU(2) symmetry on the impurity site. Finally, we add the disperson part (eq.~\ref{kineticEnergy}) back into the Hamiltonian (since it is also vanishingly small close to the Fermi surface and can therefore lead to gapless excitations). The complete Hamiltonian is
\begin{equation}\begin{aligned}
	\Delta \tilde H = \sum_{\vec q,\sigma}\left[\Delta(\vec q)\bar N_{\vec k_N + \vec q, \sigma} -2\Delta(\vec q)\bar N_{\vec k_N, \sigma}\bar N_{\vec k_N + \vec q, \sigma} -\Delta(\vec q) \left(b^\dagger_{\vec k_N, \sigma} b^\dagger_{\vec k_N, \bar\sigma} b_{\vec k_N + \vec q, \bar\sigma}b_{\vec k_N + \vec q, \sigma} + \text{h.c.}\right) \right]~,
\end{aligned}\end{equation}
where \(\Delta(q) = -{J^*(\vec q)}^2/8W\) is a positive energy scale.

@@

\section{Nature of Low-Energy Excitations in Various Regimes}
\subsection{Metallic Regime}
\paragraph{Fermi liquid excitations, \(|W| < |W|^*\):}
As discussed in Sec.~\ref{2CKEmerge}, the effective description of the impurity model for \(|W| > |W|^*\) becomes that of a 2CK. Prior to that (\(|W| < |W|^*\)), the entire Fermi surface continues to screen the impurity, and this leads to local Fermi liquid description of the impurity excitations. Using our translation operator prescription for mapping an impurity model to a lattice model (eq.~\ref{tilingPrescription}), these local Fermi liquid excitations arising from each auxiliary model coherently combine to create extended Fermi liquid excitations in the lattice model. This is therefore the nature of the metal for \(|W| < |W|^*\), and further support for this is found from the non-zero values of the quasiparticle residue in this regime (right panel of Fig.~\ref{channelDecoupling}).

\paragraph{Marginal Fermi liquid, \(|W| > |W|^*\):}
We have demonstrated in Subsection.~\ref{marginalFL} that the low-lying excitations of the pseudogap phase are described by a marginal Fermi liquid. This is linked to the fact that the excitations within the impurity model become many-particle in nature - they involve fluctuations on multiple sites. This should be contrasted with the local Fermi liquid excitations that involve only the conduction bath site next to the zeroth site. Upon tiling, we can immediately see that the resulting metal is not a Fermi liquid; the extended excitations will involve excitations of multiple \(k-\)states, and just not a simple superposition of \(1-\)particle excitations (that gave rise to the Fermi liquid). The build up of additional correlations within the marginal Fermi liquid (in comparison to the Fermi liquid) can also be ascertained from the fact that the local physics is determined by an impurity model which is highly local in \(k-\)space (due to the decoupling of the Fermi arms, see Sec.~\ref{2CKEmerge}). Such a local-in-\(k-\)space impurity model is highly non-local in real-space, consisting of long-ranged interactions between the impurity and the conduction bath. This is another signature of the non-Fermi liquid nature of the excitations obtained upon tiling.

\paragraph{At the critical point, \(|W| \lesssim |W|^*\):}

\subsection{Heisenberg model as a low-energy description of the insulator}

In the insulating phase, the ground state of each auxiliary model hosts a decoupled local moment. Upon applying the tiling procedure, the lattice model ground state becomes that of the Hubbard model in the atomic limit. In order to lift the extensive degeneracy of the state, we will now take into consideration inter-auxiliary model virtual scattering processes that were subdominant in the metallic phase and were hence ignored. These one-particle scattering processes lead to the emergence of a nearest-neighbour superexchange interaction.

For simplicity, we consider two impurity sites labelled \(1\) and \(2\) associated with two nearest-neighbour auxiliary models. The ground state subspace is four-fold degenerate:
\begin{equation}\begin{aligned}
	\ket{\Psi_L} = \left\{\ket{\sigma_1,\sigma_2}\right\}~,\quad \sigma_i = \pm 1~,
\end{aligned}\end{equation}
where \(\sigma_i\) is the spin state of site \(i\). This ground state is derived from the following "zeroth order" Hamiltonian that emerges in the local moment phase of the auxiliary models when all scattering processes between the impurity and conduction bath are RG-irrelevant:
\begin{equation}\begin{aligned}
	H_0 = -\frac{U}{2}\sum_{i=1,2}\left(n_{i \uparrow} - n_{i \downarrow}\right)^2~;
\end{aligned}\end{equation}
the local correlation on the impurity site becomes the largest scale in the problem in this phase and pushes the \(\ket{n_i=2}\) and \(\ket{n_i=0}\) states to high energies. This then defines the high-energy subspace for our calculation:
\begin{equation}\begin{aligned}
	\ket{\Psi_H} = \ket{C_1,C_2}~,
\end{aligned}\end{equation}
where \(C_i\) can take values 0 or 2, indicating that the state \(i\) is either empty or full, respectively. Both the double and hole states exist at a charge gap of the order of \(U/2\) above the low-energy singly-occupied subspace defined by the states \(\ket{\Psi_L}\).

>>>>

%! TEX program = latexmk
\documentclass[reprint,hidelinks,onecolumn]{revtex4-2}
\usepackage{braket,amsmath,color,graphicx,bookmark,bm,amssymb}
\graphicspath{{./figures/}}
\bibliographystyle{apsrev4-2}
\begin{document}

@@

\subsection{Non-Fermi liquid behaviour in the pseudogap}\label{marginalFL}
It is well-known that the two-channel Kondo model hosts non-Fermi liquid excitations at low-energies~\cite{Tsvelick_weigmann_mchannel_1985,emery_kivelson}; that this also happens in the present model is corroborated by our calculations of the quasiparticle residue \(Z_\text{imp}\) for the local gapless excitations. For a single-particle Greens function of an interacting model, the quasiparticle residue \(Z\) is defined as the spectral weight of low-energy resonance within the spectral function:
\begin{equation}\begin{aligned}
	G_k = \frac{Z}{\omega - \varepsilon_k - i \Sigma^{\prime\prime}} + G_\text{incoh}~,\quad 1/Z = 1- \frac{d \Sigma^{\prime}}{d\omega}~,
\end{aligned}\end{equation}
where \(\Sigma^\prime\) and \(\Sigma^{\prime\prime}\) are the real and imaginary parts of the self-energy, and \(G_\text{incoh}\) is contribution to the Greens function arising from incoherent short-lived excitations as well as high-energy features that are decoupled from the low-energy excitations~\cite{varma-physrevlett.63.1996}. Small values of \(Z\) indicate that the excitations cannot be described in terms of a Fermi liquid theory. Indeed, as shown in the right panel of Fig.~\ref{channelDecoupling}, the quasiparticle residue \(Z_\text{imp}\) of the impurity excitations becomes vanishingly small in the pseudogap region, signalling that the excitations are of a non-Fermi liquid nature. Given that the pseudogap phase is close to a critical point in the impurity model phase diagram (separating the local Fermi liquid and local moment phases), it is not surprising that a quantum critical model emerges at that point.

Coleman and Ioffe have shown that by mapping the two-channel Kondo problem to a ``compactified" form where the impurity spin hybridises with the spin and isospin degrees of freedom of a single conduction channel, the impurity self-energy can be shown to display marginal Fermi liquid behaviour~\cite{Coleman_tsvelik,Schofield1997}:
\begin{equation}\begin{aligned}
	\Sigma \propto \omega \ln \omega~.
\end{aligned}\end{equation}
Some of us have shown in a previous work~\cite{Patra2023MCK} that this can also be obtained from a renormalisation group calculation of the low-energy excitations of the two-channel overscreened fixed point. For the sake of completeness, we have reproduced the calculation in Appendix~\ref{mflDerivation}.

\subsection{Emergence of long-ranged interactions near the NFL-Mott insulator transition}\label{hkmEmerge}
We will now obtain a low-energy Hamiltonian for the excitations very close to the transition, where the Kondo coupling is about to turn irrelevant. In order to do this, we carry out a perturbation-theoretic treatment of the fixed point Hamiltonian in this region of the phase diagram, by considering the effects of a small fixed point Kondo scattering probability \(J^*\) in the backdrop of a larger bath interaction parameter \(W\). We note that this is in contrast to the state of affairs that gave rise to the marginal Fermi liquid excitations in Subsection ~\ref{marginalFL}, where the Kondo coupling fixed the ground state and the conduction bath hopping processes led to the low-energy excitations; it is this contrast that leads to the emergence of a new effective Hamiltonian close to the transition.

The complete Hamiltonian (details described in Appendix~\ref{hkmDerivation}) is
\begin{equation}\begin{aligned}
	\Delta \tilde H = \sum_{{\bf q}, \sigma}\text{sign}\left(\varepsilon_{{\bf N}_1 + {\bf q}}\right)\frac{\varepsilon_{{\bf N}_1 + {\bf q}}^2}{-W} r_{{\bf q},\sigma} + \sum_{{\bf q}_1,{\bf q}_2, \sigma}\frac{{J^*}^2}{-4W}\left[r_{{\bf q}_1 \sigma} \left(1 - r_{{\bf q}_2 \bar\sigma}\right) - (1 - \delta_{{\bf q}_1,{\bf q}_2})\phi^\dagger_{{\bf q}_1,\bar\sigma}\phi^\dagger_{{\bf q}_1,\sigma}\phi_{{\bf q}_2, \sigma}\phi_{{\bf q}_2, \bar\sigma}\right]~,
\end{aligned}\end{equation}
where the fermionic operator \(\phi_{{\bf q}, \sigma}\) defines emergent relative modes,
\begin{equation}\begin{aligned}
	\phi_{{\bf q}, \sigma} = \frac{1}{\sqrt 2}\left(c_{{\bf N}_1 + {\bf q},\sigma} - c_{{\bf N}_1 + {\bf Q}_1 - {\bf q}, \sigma}\right)~,
\end{aligned}\end{equation}
that are shifted by an excitation momentum \({\bf q}\) away from the nodal point \({\bf N}_1 = \left(\pi/2, \pi/2\right)\) and its nested partner \({\bf N}_1 + {\bf Q}_1 = \left(-\pi/2, -\pi/2\right)\). The number operator \(r_{{\bf q} \sigma}\) is defined as \(r_{{\bf q} \sigma} = \phi_{{\bf q}, \sigma}\dagger \phi_{{\bf q}, \sigma}\). \(\varepsilon_{{\bf N}_1 + {\bf q}}\) represents the non-interacting kinetic energy of the conduction bath.

In order to extract the singular, critical part of the Hamiltonian, we focus on the \({\bf q}_1 = {\bf q}_2\) forward scattering processes:
\begin{equation}\begin{aligned}\label{HKModel}
	\Delta \tilde H_{{\bf q}_1 = {\bf q}_2} = \sum_{{\bf q},\sigma}\epsilon_{{\bf q}}{r}_{{\bf q},\sigma} + u\sum_{{\bf q}, \sigma}r_{{\bf q} \sigma} r_{{\bf q} \bar\sigma}
\end{aligned}\end{equation}
where \(\epsilon_{\bf q}\) and \(u\) are emergent energy scales:
\begin{equation}\begin{aligned}
	\epsilon_{\bf q} = \text{sign}\left(\varepsilon_{{\bf N}_1 + {\bf q}}\right)\frac{\varepsilon_{{\bf N}_1 + {\bf q}}^2}{-W} + \frac{{J^*}^2}{-4W},~u = \frac{{J^*}^2}{4W}~.
\end{aligned}\end{equation}
Eq.~\ref{HKModel} describes a Hatsugai-Kohmoto model~\cite{Baskaran1991,Hatsugai1992}, an exactly solvable model of correlated electrons; by tuning the ratio of the interaction strength \(u\) and the bandwidth \(w\), the model describes a transition between a non-Fermi liquid and a Mott insulator. The non-Fermi liquid nature of the gapless excitations in the metallic phase of the model manifests in the form of a divergent one-particle self-energy at the non-interacting Fermi surface~\cite{Phillips2020}:
\begin{equation}\begin{aligned}
	\Sigma_{{\bf q}}(\omega) = -\frac{u^2/4}{\omega - \epsilon_{\bf q}}, ~\Sigma_{\epsilon_{\bf q}=0}(\omega \to 0)  \to \infty~.
\end{aligned}\end{equation}
Beyond the transition and inside the Mott insulating phase, this zero frequency pole continues to exist and marks the hard gap in the spectral function for charge excitations.

@@

\section{Nature of Low-Energy Excitations in Various Regimes}
\subsection{Fermi liquid excitations, \(|W| < |W|^*\)}
As discussed in Sec.~\ref{2CKEmerge}, the effective description of the impurity model for \(|W| > |W|^*\) becomes that of a two-channel Kondo (2CK) model. Prior to that (\(|W| < |W|^*\)), the entire Fermi surface continues to screen the impurity, and this leads to local Fermi liquid description of the impurity excitations. Using our translation operator prescription for mapping an impurity model to a lattice model (eq.~\ref{tilingPrescription}), these local Fermi liquid excitations arising from each auxiliary model coherently combine to create extended Fermi liquid excitations in the lattice model. This is therefore the nature of the metal for \(|W| < |W|^*\), and further support for this is found from the non-zero values of the quasiparticle residue in this regime (right panel of Fig.~\ref{channelDecoupling}).

\subsection{Marginal Fermi liquid, \(|W| > |W|^*\)}
We have demonstrated in Subsection.~\ref{marginalFL} that the low-lying excitations of the pseudogap phase are described by a marginal Fermi liquid. This is linked to the fact that the excitations within the impurity model become many-particle in nature - they involve fluctuations on multiple sites. This should be contrasted with the local Fermi liquid excitations that involve only the conduction bath site next to the zeroth site. Upon tiling, we can immediately see that the resulting metal is not a Fermi liquid; the extended excitations will involve excitations of multiple \(k-\)states, and just not a simple superposition of \(1-\)particle excitations (that gave rise to the Fermi liquid). The build up of additional correlations within the marginal Fermi liquid (in comparison to the Fermi liquid) can also be ascertained from the fact that the local physics is determined by an impurity model which is highly local in \(k-\)space (due to the decoupling of the Fermi arms, see Sec.~\ref{2CKEmerge}). Such a local-in-\(k-\)space impurity model is highly non-local in real-space, consisting of long-ranged interactions between the impurity and the conduction bath. This is another signature of the non-Fermi liquid nature of the excitations obtained upon tiling.

\subsection{At the critical point, \(|W| \lesssim |W|^*\)}
Crucially, we find that during the passage through the pseudogap phase, the nature of the non-Fermi liquid metal itself morphs; the marginal Fermi liquid description is found to break down at the transition, and is replaced by the non-Fermi liquid excitations of a Hatsugai-Kohmoto model. The singular nature of the excitations in this regime can be ascertained from the self-energy displaying divergent behaviour (see Subsection~\ref{hkmEmerge}). This can be attributed to the extremely non-local form of the impurity model (arising from the fact that it couples to only the nodal points in \(k-\)space). This divergent self-energy is consistent with our numerical calculations of the impurity self-energy which showed that the self-energy poles approach the \(\omega=0\) point as the system is tuned towards the transition.

\subsection{Heisenberg model as a low-energy description of the Mott insulator}

In the insulating phase, the ground state of each auxiliary model hosts a decoupled local moment. Upon applying the tiling procedure, the lattice model ground state becomes that of the Hubbard model in the atomic limit. In order to lift the extensive degeneracy of the state, we will now take into consideration inter-auxiliary model virtual scattering processes that were subdominant in the metallic phase and were hence ignored. These one-particle scattering processes lead to the emergence of a nearest-neighbour superexchange interaction. The calculation is a straightforward application of second order perturbation theory (Schrieffer-Wolff transformation). In order to allow virtual fluctuations that can lift the large ground state degeneracy and lower the energy, we consider (perturbatively) the effects of an irrelevant single-particle hybridisation that connects the nearest-neighbour sites. This perturbation Hamiltonian is therefore of the form
\begin{equation}\begin{aligned}
	H_t = \sum_\omega V(\omega) \mathcal{P}(\omega)~\sum_\sigma\left(c^\dagger_{1\sigma}c_{2\sigma} + \text{h.c.}\right) ~,
\end{aligned}\end{equation}
where \(V(\omega)\) only acts on states at the energy scale \(\omega\); the renormalisation of \(V\) is encoded in the fact that \(V(\omega)\) is largest for the excited states and vanishes at low-energies: \(V(\omega \to 0) = 0\).

We show the details in Appendix~\ref{heisenbergDerivation}. The final result is the emergence of the following effective Hamiltonian for the low-energy dynamics between any two nearst-neighbour sites \(i\) and \(j\):
\begin{equation}\begin{aligned}
	H_\text{eff} = 2\frac{V_H^2}{U/2}\left(2S_i^z S_j^z + S_i^+S_j^- + S_i^-S_j^+\right) = J_\text{eff} {\bf S}_i\cdot{\bf S}_j~,
\end{aligned}\end{equation}
where the effective antiferromagnetic Heisenberg coupling is \(J_\text{eff} = \frac{8V_H^2}{U}\) and \(V_H \equiv V(\omega \to U/2)\) is the impurity-bath hybridisation at energy scales of the order of the Mott gap, in the sense of an RG flow.

\clearpage
\section*{Appendices}
\appendix
\subsection{Marginal Fermi liquid excitations in a two-channel Kondo model}\label{mflDerivation}
By considering minimal hopping into the conduction bath and treating the effects of this hopping perturbatively against the overscreened screened ground state (determined by an intermediate coupling \(J^*\)), the low-energy excitations above the ground state of the two-channel Kondo (2CK) model are described by the following Hamiltonian~\cite{Patra2023MCK}:
\begin{eqnarray}
	\label{nfl_terms}
	H_\mathrm{eff} =& \frac{8t^2}{{J}}\left[\left(S^z_{0,1}\right)^2 S^+_{0,2} + \left(S^z_{0,2}\right)^2 S^+_{0,1}\right] \left(S^-_{1,1} + S^-_{1,2}\right) + \mathrm{h.c.};
\end{eqnarray}
where \({\boldsymbol S}_{i,l} = \frac{1}{2}\sum_{\sigma_1,\sigma_2}c^\dagger_{i,\sigma_1,l}{\boldsymbol \sigma}_{\sigma_1,\sigma_2}c_{i,\sigma_2,l}\) are spin operators for the \(i^\text{th}\) site on the \(l^\text{th}\) conduction channel (\(l\) can be 1 or 2 for a two-channel model). The labels \(\sigma_1\) and \(\sigma_2\) are spin indices that can be \(\pm 1\). The site index \(0\) represents the bath sites connected immediately to the impurity site. The operator \(c^\dagger_{i,\sigma,l}\) creates an electron with spin \(\sigma\) on the \(i^\text{th}\) site of the \(l^\text{th}\) conduction channel.

\subsection{Emergence of Hatsugai-Kohmoto model near the NFL-Mott insulator transition}\label{hkmDerivation}
As the bath interaction \(W\) is tuned through the L-PG phase, the nodal region is the last to decouple from the impurity. This allows us to write down a simpler Kondo model near the transition, where only the nodal region is hybridising with the impurity spin through Kondo interactions. This is done by retaining only those scattering processes \({\bf k}_1 \to {\bf k}_2\) that originate from and end at \(k-\)points within a small neighborhood of width \(|{\bf q}|\) around the four nodal points: \({\bf k}_1, {\bf k}_2 \in {\bf N} + {\bf q}, |{\bf q}| \ll \pi\), where \({\bf N}\) can be any one of the four nodal points \({\bf N}_{1} = \left(\pi/2, \pi/2\right), {\bf N}_{2} = \left(-\pi/2, \pi/2\right)\) and \({\bf N}_1 + {\bf Q}_1\) and \({\bf N}_{2} + {\bf Q}_2\), where \({\bf Q}_1 = \left(-\pi,-\pi \right)\) and \({\bf Q}_2 = \left( \pi, -\pi \right) \) are the two nesting vectors. We have already seen, in Subsection ~\ref{2CKEmerge}, that the pseudogap regime displays a decoupling of the Brillouin zone into sets of points that have no explicit connecting term in the Hamiltonian. Near the transition, the neighbourhoods of \({\bf N}_{1}\) and \({\bf N}_{1} + {\bf Q}_1\) form one such connected set, while the other two nodes \({\bf N}_{2}\) and \({\bf N}_{2} + {\bf Q}_2\) form another connected set, but these two sets are disconnected from each other, in the sense of a two-channel model. We also assume that the window of \({\bf q}\) is small enough so that the fixed point Kondo coupling values \(J^*({\bf q}_1, {\bf q}_2)\) for the scattering processes involving \({\bf q}_1\) and \({\bf q}_2\) can be replaced by an average value \(J^*\).

We find that the sum degrees of freedom \(s_{{\bf q},\sigma}\) do not enter the Kondo scattering Hamiltonian \(\tilde H_\text{imp-cbath}\) (for the single channel \(l=1\)), and it is expressed purely in terms of the relative degrees of freedom:
\begin{equation}\begin{aligned}
	\tilde H_\text{imp-cbath} = J^*\sum_{{\bf q}_1, {\bf q}_2,\alpha,\beta}{\bf S}_d\cdot{\boldsymbol\sigma}_{\alpha\beta} {\phi}^\dagger_{{\bf q}_1\alpha} \phi_{{\bf q}_2\beta}~.
\end{aligned}\end{equation}
We now proceed to rewrite the other parts of the Hamiltonian in terms of these new degrees of freedom. The kinetic energy can be written as
\begin{equation}\begin{aligned}\label{kineticEnergy}
	\tilde H_\text{cbath} &= \sum_{{\bf q},\sigma}\varepsilon_{{\bf N}_1 + {\bf q}}\left(n_{{\bf N}_1 + {\bf q},\sigma} + n_{{\bf N}_1 + {\bf Q}_1 - {\bf q},\sigma}\right)\\
						  &= \sum_{{\bf q},\sigma}\varepsilon_{{\bf N}_1 + {\bf q}}\left({\psi}^\dagger_{{\bf q},\sigma} {\phi}_{{\bf q},\sigma} + {\phi}^\dagger_{{\bf q},\sigma} \psi_{{\bf q},\sigma}\right)~,
\end{aligned}\end{equation}
We also consider a simplified form of the bath interaction (for the channel \(l=1\)), taking into account the density-density correlations in \(k-\)space:
\begin{widetext}
\begin{equation}\begin{aligned}
	\tilde H_\text{cbath-int} =& -W\sum_{{\bf q},\sigma}\left(n_{{\bf N}_1 + {\bf q},\sigma} + n_{{\bf N}_1 + {\bf Q}_1 - {\bf q},\sigma}\right) + W\sum_{{\bf q}_1,{\bf q}_2}\left(n_{{\bf N}_{1} + {\bf q}_1, \uparrow} + n_{{\bf N}_{1} + {\bf Q}_{1} - {\bf q}_1, \uparrow} \right) \left( n_{{\bf N}_{1} + {\bf q}_2, \downarrow} + n_{{\bf N}_{1} + {\bf Q}_{1} - {\bf q}_2, \downarrow} \right)~,\\
	=& -W\sum_{{\bf q},\sigma}\left(s_{{\bf q},\sigma} + r_{{\bf q},\sigma}\right) + W\sum_{{\bf q}_1,{\bf q}_2}\left(s_{{\bf q}_1,\uparrow} + r_{{\bf q}_1,\uparrow}\right) \left(s_{{\bf q}_2,\downarrow} + r_{{\bf q}_2,\downarrow}\right)~.
\end{aligned}\end{equation}
\end{widetext}
We used eq.~\ref{numberOperatorRelation} in order to obtain the final form.
